---
title: "ST558_HW3 - Lee Worthington"
format: html
editor: visual
---

# -------------------------------------------
# Task 1: conceptual questions
# -------------------------------------------

#### Question 1 - If your working directory is myfolder/homework/, what relative path would you specify to get the file located at myfolder/MyData.csv?
> If the working directory is myfolder/homework/ then to get the file MyData.csv from myfolder the relative path would be '../MyData.csv'

#### Question 2 - What are the major benefits of using R projects?
> Rprojects gives you an easy way to:
>
> - Set a structured working directory and workspace with a specific environment for all your relevant files and objects
> - Implement version control using git
> - Provide reproducibility as the saved environment allows others to generate the same results
> - Allow easier collaboration with github

#### Question 3 - What is git and what is github?
> Git is version control software, whereas github is a hosting service for git based projects

#### Question 4 - what are the two main differences between a tiblle and a data.frame?
> data.frame vs tibble
>
> - When printing tibbles print additional information such as the number of rows and columns
> - Tibbles don’t coerce down to a vector when you subset to a single column using [], instead they return a tibble, unlike a data frame which coerces down to a vector

#### Question 5 - rewrite the following nested function call using baseR's chaining operator
```{r}
#| eval: false

# Original code 
arrange(filter(select(as_tibble(iris), starts_with("Petal"), Species), Petal.Length < 1.55), Species)

#New code
iris |>
  as_tibble() |>
  select(starts_with("Petal"), Species) |>
  filter(Petal.Length < 1.55) |>
  arrange(Species)
```

#### Question 6 - what is meant by long format data and wide format data? Which is generally prefered for statistical analysis?
> In long format each row represents a single observation and each column represents a single variable, whereas in wide format you may have multiple observations in a single row.
  

# -------------------------------------------
# Task 2: reading delimited data
# -------------------------------------------

#### Glass data
```{r}
#| eval: true

# load library
library(tidyverse)

# read in file
glass_data <- read_csv(
  'https://www4.stat.ncsu.edu/~online/datasets/glass.data',
  col_names = c('id', 'ri', 'na', 'mg', 'al', 'si', 'k', 'ca', 'ba', 'fe', 'type_of_glass')
)

# call tibble
glass_data

# cstart chain on glass_data
glass_data |>
  
  # update type_of_glass field
  mutate(type_of_glass = ifelse(type_of_glass == 1, 'building_windows_float_processed',
                ifelse(type_of_glass == 2, 'building_windows_non_float_processed',
                       ifelse(type_of_glass == 3, 'vehicle_windows_float_processed',
                              ifelse(type_of_glass == 4, 'vehicle_windows_non_float_processed',
                                     ifelse(type_of_glass == 5, 'containers',
                                            ifelse(type_of_glass == 6, 'tableware',
                                                   ifelse(type_of_glass == 7, 'headlamps', NA)))))))) |>
  
  # filter on fe variable (& or , should work)
  filter(fe < 0.2 & type_of_glass %in% c('tableware', 'headlamps'))

```

#### Yeast data
```{r}
#| eval: true

# read in file (this data is separated by spaces (https://readr.tidyverse.org/reference/read_table.html)
yeast_data <- read_table(
  'https://www4.stat.ncsu.edu/~online/datasets/yeast.data',
  col_names = c('seq_name', 'mcg', 'gvh', 'alm', 'mit', 'erl', 'pox', 'vac', 'nuc', 'class')
)

# call tibble
yeast_data 

# apply chain
yeast_data |>
  # filter data
  select(-seq_name, -nuc) |>
  # group by class
  group_by(class) |>
  # create new columns using mutate with summary stats using across on numeric columns
  mutate(across(where(is.numeric), mean, .names = 'mea_{.col}')) |>
  mutate(across(where(is.numeric), median, .names = 'median_{.col}'))
  
```

# -------------------------------------------
# Task 2: combining excel and delimited data
# -------------------------------------------

#### Part 1
```{r}
#| eval: true

# load library
library(readxl)

# load data
white_wine <- read_excel('white-wine.xlsx')

# print tibble
white_wine
  
```

#### Part 2
```{r}
#| eval: true

# load variable names into a tibble
field_names <- read_excel(
  'white-wine.xlsx',
  sheet = excel_sheets('white-wine.xlsx')[2]
)
  
# update white_wine fuse field_names vector
colnames(white_wine) <- field_names[[1]]

# print results
white_wine

```

#### Part 3
```{r}
#| eval: true

# add new column to white_wine
white_wine <- white_wine |>
  # add new field
  mutate(type = 'white')

```

#### Part 4
```{r}
#| eval: true

# read data
red_wine <- read_delim(
  'https://www4.stat.ncsu.edu/~online/datasets/red-wine.csv',
  delim = ';'
)

# update red_wine fuse field_names vector
colnames(red_wine) <- field_names[[1]]

# add new column to white_wine
red_wine <- red_wine |>
  mutate(type = 'red')

red_wine

```

#### Part 5
```{r}
#| eval: true

# part 5 combine white and red, ?dplyr::bind_rows
wine <- dplyr::bind_rows(white_wine, red_wine)

# part 6
wine |>
  # filter on quality 6
  filter(quality > 6.5 & alcohol < 132) |>
  # sort rows by descending quality 7
  arrange(desc(quality)) |>
  # select specified fields
  select(contains('acid'), alcohol, type, quality) |>
  # mean and stddev for alcohol by quality
  group_by(quality) |>
  mutate(alcohol_mean = mean(alcohol), alcohol_stddev = sd(alcohol))
```


# -------------------------------------------
# Task 3: database practice
# -------------------------------------------

```{r}
#| eval: true

# install library
#install.packages('DBI')
#install.packages('RSQLite')

# load libraries
library(DBI)
library(RSQLite)

# 1 connect to db
con <- dbConnect(RSQLite::SQLite(), 'lahman.db')
dbListTables(con)

# 2 load data to a tibble
teams_data <- tbl(con, 'Teams')

teams_data |>
  filter(yearID == 2015)

# 3 use sql to do the same thing
tbl(con, 
    sql('
    select *
    from Teams t
    where 1=1
    and yearID = 2015
    '
))

# 4 return all players in the hall of fame, the year, and the category
hall_of_fame <- tbl(con, 
    sql('
    select category, playerId, yearID 
    from HallOfFame
    where 1=1
    and category = "Player"
    '
))
hall_of_fame

# 5 combine previous data wiht the People table
hof_players <- tbl(con, 
    sql('
    select 
    hof.category,
    
    hof.playerId, 
    p.nameFirst,
    p.nameLast,
    
    hof.yearID 

    from HallOfFame hof
    inner join People p
    on 1=1
    and hof.playerID = p.playerID
    
    where 1=1
    and hof.category = "Player"
    '
))

# 6 Managers table manipulation
managers <- tbl(con,'Managers') |>
  # select specified columns
  select(playerID, G, W, L) |>
  # grooup by playerID
  group_by(playerID) |>
  # summary stats
  summarize(
    # total games managed
    G_managed = sum(G, na.rm = TRUE),
    #total wins
    Total_W = sum(W, na.rm = TRUE),
    # total losses
    Total_L = sum(L, na.rm = TRUE)
  ) |>
  # execute up to this point to calculate summary stats
  collect() |>
  # calculate win/loss percentage in a new field
  mutate(win_percentage = Total_W/G_managed) |>
  #mutate(loss_percentage = Total_L/G_managed) |>
  # sort by win percentage
  arrange(desc(win_percentage))

# 7 combine results
# pull all hof data, regardless of category
hof <- tbl(con, 
    sql('
    select 
    hof.category,
    
    hof.playerID, 
    p.nameFirst,
    p.nameLast,
    
    hof.yearID 

    from HallOfFame hof
    inner join People p
    on 1=1
    and hof.playerID = p.playerID
    
    where 1=1
    '
)) |>
  collect()

# combine the data
combined_data <- inner_join(
  managers,
  hof,
  by = c("playerID" = "playerID")
)

```


```








